Template \aastex Article with Examples: v6.3\footnote{Released on June, 10th, 2019}

### 2020-04-29 writeups/Su_disk_disp.tex
final Dong review

%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%%        wider left and right margins. This uses the Daniel
%%        Foreman-Mackey and David Hogg design.
%%  RNAAS       : Preferred style for Research Notes which are by design
%%                lacking an abstract and brief. DO NOT use \begin{abstract}
%%                and \end{abstract} with this style.
    \documentclass[twocolumn]{aastex63}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        graphicx,
        ae, aecompl,
        booktabs,
        caption,
        subcaption,
        xcolor,
    }
    \usepackage[T1]{fontenc}
    \captionsetup{compatibility=false}

@@

Large planetary spin-orbit misalignments (obliquities) may strongly influence
atmospheric circulation and tidal heating in the planet. A promising avenue to
generate obliquities is via spin-orbit resonances, where the spin and orbital
precession frequencies of the planet cross each other as the system evolves in
time. One such mechanism involves a dissipating (mass-losing) protoplanetary
disk driving orbital precession of an interior planet. We study this scenario
analytically in this paper, and obtain the mapping between the initial spin
orientation and the final obliquity. We show that (i) under adiabatic evolution
(i.e.\ the disk dissipates at a sufficiently slow rate), the final planetary
obliquities bifurcate into distinct tracks governed by interactions with the
resonance; (ii) under nonadiabatic evolution, a broad range of obliquities can
be excited. We obtain analytical expressions for the final obliquities for most
interesting regimes of parameter space. The dynamical system studied in this
paper is an example of ``Colombo's top'', and our analysis and results can be
adapted for other applications.

@@

@@

A rotating planet is subjected to gravitational torque from its host star,
making its spin axis precess around its orbital (angular momentum) axis.  Now
suppose the orbital axis precesses around another fixed axis---such orbital
precession could arise from gravitational interactions with other masses in the
system (e.g.\ planets, external discs, or binary stellar companion). What is the
dynamics of the planetary spin axis?  How does the spin axis evolve as the spin
precession rate, the orbital precession rate, or their ratio, gradually changes
in time?

\citet{colombo1966} was the first to point out the importance of the above
simple model in the study of the obliquity (the angle between the spin and
orbital axes) of planets and satellites. Subsequent works
\citep{peale1969,peale1974possible,ward1975tidal,henrard1987} have revealed rich
dynamics of this model. With appropriate modification, this model can be used as
a basis for understanding the evolution of rotation axes of celestial bodies.
Indeed, many contemporary problems in planetary/exoplanetary dynamics can be
cast into a form analogous to this simple model or its variants
\citep[e.g.][]{ward2004I, fabrycky_otides, batygin2013magnetic, lai2014star,
anderson2018teeter}.

In this paper we present a systematic investigation on the secular evolution of
Colombo's top, starting from general initial conditions. Our study includes
several new analytical results that go beyond previous works. While our results
are general, we frame our study in the context of generating exoplanet
obliquities from planet-disc interaction with a dissipating disc.

@@

The Millholland \& Batygin study assumes a negligible initial planetary
obliquity. This assumption is intuitive, since the planet attains its spin
angular momentum from the disk. But it may not always be satisfied. In
particular, the formation of rocky planets through planetesimal accretion can
lead to a wide range of obliquities, especially if the final spin is imparted by
a few large bodies \citep{dones1993does, lissauer1997accretion,
miguel2010planet}. Such ``stochastic'' accretion likely happened for terrestrial
planets in the Solar System. Giant impacts may have also played a role in the
formation of the close-in multiple-planet systems discovered by the Kepler
satellite \citep[e.g.][]{inamdar2015formation, izidoro2017breaking}.

@@

The paper is organized as follows. In Section~\ref{s:eq}, we review the relevant
spin-orbit dynamics and key concepts that are used in the remainder of the
paper. In Sections~\ref{s:ad} and~\ref{s:nonad}, we study the evolution of the
system when the disk dissipates on different timescales, from highly adiabatic
to nonadiabatic. Analytical results are presented to explain the numerical
results in both limits. We discuss the implications of our results in
Section~\ref{s:disc}. Our primary physical results consist of
Fig.~\ref{fig:ad_ensemble} in the adiabatic limit and
Fig.~\ref{fig:3_ensemble_05_15} in the nonadiabatic limit. Some detailed
calculations are relegated to the appendices, including a leading-order
estimate of the final planetary obliquities given small initial spin-disk
misalignment angles in Appendix~\ref{s:ad_approx}.

@@

The spin axis $\uv{s}$ of the planet tends to precess around its orbital
(angular momentum) axis $\uv{l}$, driven by the gravitational torque from the
host star acting on the planet's rotational bulge. On the other hand, $\uv{l}$
and the disc axis $\uv{l}_{\rm d}$ precess around each other due to
gravitational interactions. We assume $S \ll L \ll L_{\rm d}$, so $\uv{l}_{\rm
d}$ is nearly constant and $\uv{l}$ experiences negligible backreaction torque
from $\uv{s}$. The
equations of motion for $\uv{s}$ and $\uv{l}$ are
\begin{align}
    \rd{\uv{s}}{t} &= \omega_{\rm sl} \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        \equiv \alpha \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}},\label{eq:dsdt}\\
    \rd{\uv{l}}{t} &= \omega_{\rm ld}\p{\uv{l} \cdot \uv{l}_{\rm d}}
            \p{\uv{l} \times \uv{l}_{\rm d}}
        \equiv -g\p{\uv{l} \times \uv{l}_{\rm d}},
            \label{eq:dldt}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &\equiv \frac{3GJ_2 M_pR_p^2 M_\star}{2a_p^3 I_p\Omega_p}
        = \frac{3k_{\rm qp}}{2k_{\rm p}} \frac{M_\star}{m_{\rm p}}
            \p{\frac{R_{\rm p}}{a_{\rm p}}}^3 \Omega_{\rm p},\label{eq:wsl}\\
    \omega_{\rm ld} &\equiv \frac{3M_{\rm d}}{4M_\star}\p{\frac{a_{\rm
            p}}{r_{\rm d}}}^3 n .\label{eq:wld}
\end{align}
In Eq.~\eqref{eq:wsl}, $I_p = k_p M_pR_p^2$ (with $k_p$ a constant) is the
moment of inertia and $J_2=k_{\rm qp}\Omega_p^2 (R_p^3/GM_p)$ (with $k_{qp}$ a
constant) the rotation-induced (dimensionless) quadrupole of the planet [for a
body with uniform density, $k_p=0.4, k_{qp}=0.5$; for giant planets, $k_p\simeq
0.25$ and $k_{qp}\simeq 0.17$ \citep[e.g.][]{lainey2016quantification}]. In
Eq.~\eqref{eq:wld}, $n \equiv \sqrt{GM_\star/a_{\rm p}^3}$ is the planet's
orbital mean motion, and we have assumed $r_d\gg a_p$ and included only the
leading-order (quadrupole) interaction between the planet and disc. We define
three relative inclination angles via
\begin{equation}
  \uv{s} \cdot \uv{l}\equiv \cos\theta,\quad
  \uv{s} \cdot \uv{l}_{\rm d}\equiv \cos\theta_{\rm sd},\quad
  \uv{l} \cdot \uv{l}_{\rm d}\equiv \cos I.
\end{equation}
In our model, $I$ is a constant.
Following standard notation \citep[e.g.][]{colombo1966,peale1969,ward2004I}, we
have defined $\alpha \equiv \omega_{\rm sl}$ and $g \equiv -\omega_{\rm ld}\cos I$.

@@

In this section, we study the evolution of the planetary obliquity $\theta$ when
the parameter $\eta$ [or the disk mass $M_{\rm d}$; see Eqs.~\eqref{eq:eta}
and~\eqref{eq:deta_dt}] decreases sufficiently slowly that the evolution is
adiabatic. \textcolor{Corr}{Intuitively, this requires the disk evolution time
$t_{\rm d}$ [Eq.~\eqref{eq:dmd_dt}] be larger than the spin precession period,
$2\pi / \alpha$, i.e.\ $\epsilon = 1 / (\alpha t_{\rm d}) \lesssim 1/ (2\pi)$.}

Strictly speaking, \textcolor{Corr}{adiabaticity} requires $t_{\rm d}$ be much
larger than all timescales of the dynamical system governed by the Hamiltonian
[Eq.~\eqref{eq:H}]. This is of course not possible in all cases, as the motion
along the separatrix has an infinite period. In practice, as $\eta$ evolves, the
system only crosses the separatrix once or twice, while it spends many orbits
inside one of the three zones and far from the separatrix. Thus, a \emph{weak
adiabaticity criterion} is that, for all equilibria/fixed points, the local
circulation/libration periods are much shorter than the
\textcolor{Corr}{timescale for the} motion of the equilibria
\textcolor{Corr}{due to changing $\eta$}. If this criterion is satisfied, then
the system will evolve adiabatically for most of its
evolution save one or two separatrix crossings.

As shown in Appendix~\ref{ss:eigens}, libration about CS2 is slower than that
about CS1 or CS3. As such, it has the smallest characteristic frequency in the
system. The weak adiabaticity criterion is equivalent to
requiring that, at all times other than separatrix crossing, the obliquity of
CS2 ($\theta_2$) evolve over a longer timescale than the local libration period
about CS2, i.e.
\begin{equation}
    \abs{\rd{\theta_2}{\tau}} \lesssim
        \frac{\omega_{\rm lib}}{2\pi},\label{eq:ad_base}
\end{equation}
where
\begin{equation}
    \omega_{\rm lib} = \sqrt{\eta\sin I \sin \theta_2
        \p{1 + \eta \sin I \csc^3 \theta_2}},\label{eq:w_lib}
\end{equation}
is the libration frequency about CS2 \textcolor{Corr}{for a given $\eta$}
(Appendix~\ref{ss:eigens}). This formula differs from that given in
\citet{millholland_disk}, where the $\csc^3\!\theta_2$ term is neglected and the
square root is missing\footnote{The missing $\csc^3\!\theta_2$ term can be
traced to a $\theta \gg I$ approximation made in Eq.~(3) of \citet{ward2004II}.
Since $\theta_2 \sim I$ for $\eta \gtrsim 1$ (Fig.~\ref{fig:cs_locs}), this
approximation is not always valid.}. Differentiating Eq.~\eqref{eq:cs_zero}
gives
\begin{equation}
    \rd{\theta_2}{\tau} = -\epsilon \frac{\eta \sin\p{\theta_2 - I}}{
        \cos\p{2\theta_2} - \eta \cos\p{\theta_2 - I}},\label{eq:dq2dt}
\end{equation}
where $\epsilon = -\rdil{(\ln \eta)}{\tau}$ [Eq.~\eqref{eq:eps_def}].
\textcolor{Corr}{Eq.~\eqref{eq:ad_base} is most constraining at $\eta \sim 1$,
i.e.\ it is satisfied for all $\eta$ if it is satisfied near $\eta \sim 1$,
$\abs{\rdil{\theta_2}{\tau}} \sim \epsilon$}. Thus, weak adiabiticity requires
\textcolor{Corr}{
\begin{equation}
    \epsilon \lesssim \epsilon_{\rm c}
        \equiv \p{\frac{\omega_{\rm lib}}{2\pi}}_{\eta = 1}
            % =\frac{1}{2\pi}\sqrt{\sin I \sin \theta_{\rm 2c}
            %     \p{1 + \sin I \csc^3 \theta_{\rm 2c}}}
            \sim 0.0433,\label{eq:ad_constr}
\end{equation}
where we have used $\sin \theta_{\rm 2} \sim 1/2$ when $\eta =
1$ (for $I = 5^\circ$ and $\eta = 1$, $\theta_{\rm 2} \approx 31^\circ$)}. Since
our criterion is only a weak condition for adiabaticity, we use $\epsilon = 3
\times 10^{-4}$ in our ``adiabatic'' calculations below. We explore the
consequences of nonadiabatic evolution in Section~\ref{s:nonad}.

@@

To obtain the distribution of the final obliquities $\theta_{\rm f}$, we evenly
sample $101$ values of $\theta_{\rm sd, i}$, and for each $\theta_{\rm sd, i}$
value, we pick $101$ evenly spaced orientations of $\uv{s}$ approximately from
the ring of initial conditions having angular distance $\theta_{\rm sd, i}$ to
$\uv{l}_{\rm d}$%
%
\footnote{The actual procedure we adopt to choose the initial conditions is the
natural extension of this description to finite $\eta_{\rm i}$. Note that the
center of libration of $\uv{s}$ is CS2, which, since $\eta_{\rm i}$ is finite,
is different from $\uv{l}_{\rm d}$. Furthermore, the libration is not exactly
circular. As a result, the libration trajectories for initial conditions on the
circular ring of points having angular distance $\theta_{\rm sd, i}$ from
$\uv{l}_{\rm d}$ are not the same and will each enclose slightly different
initial phase space areas $A_{\rm i}$. Since our analytical theory assumes exact
conservation of the initially enclosed phase space area $A_{\rm i}$ for each
$\theta_{\rm sd, i}$ (see Section~\ref{ss:zone_transitions}), this discrepancy
introduces an extra deviation from the analytical prediction. To guarantee all
points for a particular $\theta_{\rm sd, i}$ have the same $A_{\rm i}$, we
instead choose initial conditions on the libration cycle going through
$\p{\theta_2 + \theta_{\rm sd, i}, \phi_2}$ [where $\p{\theta_2, \phi_2}$ are
the coordinates of CS2]. This ensures that all initial conditions for a given
$\theta_{\rm sd, i}$ enclose the same initial $A_{\rm i}$. As $\eta_{\rm i} \to
\infty$, this procedure generates initial conditions on the ring having angular
distance $\theta_{\rm sd, i}$ to $\uv{l}_{\rm d}$, recovering the procedure
given in the text.}.
%
To be concrete, we choose $\eta_{\rm i} = 10\eta_{\rm c}$ where $\eta_{\rm c}$
is given by Eq.~\eqref{eq:etac} and evolve Eqs.~\eqref{eq:dsdt_base}
and~\eqref{eq:deta_dt} until $\eta$ reaches its final value $10^{-5}$. At such a
small $\eta$, $\uv{s}$ is strongly coupled to $\uv{l}$ and the final obliquity
$\theta_{\rm f}$ is frozen. The mapping between $\theta_{\rm sd, i}$ and
$\theta_{\rm f}$ is our primary result, and is shown for $I = 5^\circ$,
$10^\circ$, and $20^\circ$ in
Figs.~\ref{fig:ad_ensemble},~\ref{fig:3_ensemble_10_35},
and~\ref{fig:3_ensemble_20_35} respectively. The blue dots represent the results
of the numerical calculation. The colored tracks are calculated
semi-analytically using the method discussed in the following subsection.

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. We adopt
convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
This definition of $A$ has two advantages: (i) it is continuous across
transitions from circulating to librating that cross the North pole ($\cos
\theta = 1$), and (ii) the areas of the three zones are equal in absolute value
to the expressions given in Eqs.~\eqref{se:area_ward}. The path over which the
integral is taken is either a libration or circulation cycle. When $\eta_{\rm i}
\gg 1$, trajectories librate about $\uv{l}_{\rm d}$ with constant $\theta_{\rm
sd}$, meaning they enclose initial phase space area
\begin{equation}
    A_{\rm i} = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_{\rm i}$, as trajectories near
CS2 or CS3 librate about these equilibria, rather than $\uv{l}_{\rm d}$, and
Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice, Eq.~\eqref{eq:ai_qsd}
holds very well when defining $\theta_{\rm sd, i}$ as the angular distance to
CS2; an exception is discussed in Section~\ref{sss:evol_traj}.

@@

The above calculation applies for a specific initial $\theta_{\rm i}$, but, as
discussed at the beginning of Section~\ref{ss:ad_ensemble}, the initial spin
orientation is more appropriately described by $\theta_{\rm sd, i}$ since
$\eta_i\gg 1$. The correct way to predict the final obliquity for a given
$\theta_{\rm sd, i}$ using Eq.~\eqref{eq:thetaf} is somewhat subtle but yields
good agreement with numerical results.

@@

We thank Kassandra Anderson for discussion and assistance in the early phase of
this work. DL thanks the Dept.\ of Astronomy and the Miller Institute for Basic
Science at UC Berkeley for hospitality while part of this work was carried out.
This work has been supported in part by the NSF grant AST-17152 and NASA grant
80NSSC19K0444. YS is supported by the NASA FINESST grant
19-ASTRO19-0041.% chktex 8

>>>>

%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%%        wider left and right margins. This uses the Daniel
%%        Foreman-Mackey and David Hogg design.
%%  RNAAS       : Preferred style for Research Notes which are by design
%%                lacking an abstract and brief. DO NOT use \begin{abstract}
%%                and \end{abstract} with this style.
    \documentclass[twocolumn,twocolappendix]{aastex63}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        graphicx,
        ae, aecompl,
        booktabs,
        caption,
        subcaption,
        xcolor,
    }
    \usepackage[T1]{fontenc}
    \captionsetup{compatibility=false}

@@

Large planetary spin-orbit misalignments (obliquities) may strongly influence
atmospheric circulation and tidal heating in the planet. A promising avenue to
generate obliquities is via spin-orbit resonances, where the spin and orbital
precession frequencies of the planet cross each other as the system evolves in
time. One such mechanism involves a dissipating (mass-losing) protoplanetary
disk that drives orbital precession of an interior planet. We study this
scenario analytically in this paper, and obtain the mapping between the general
initial spin orientation and the final obliquity. We show that (i) under
adiabatic evolution (i.e.\ the disk dissipates at a sufficiently slow rate), the
final planetary obliquity as a function of the initial spin orientation
bifurcates into distinct tracks governed by interactions with the resonance;
(ii) under nonadiabatic evolution, a broad range of obliquities can be excited.
We obtain analytical expressions for the final obliquities for various
regimes of parameter space. The dynamical system studied in this paper is an
example of ``Colombo's top'', and our analysis and results can be adapted to
other applications.

@@

\keywords{planet---star interactions}

@@

A rotating planet is subjected to gravitational torque from its host star,
making its spin axis precess around its orbital (angular momentum) axis.  Now
suppose the orbital axis precesses around another fixed axis---such orbital
precession could arise from gravitational interactions with other masses in the
system (e.g.\ planets, external disks, or binary stellar companion). What is the
dynamics of the planetary spin axis?  How does the spin axis evolve as the spin
precession rate, the orbital precession rate, or their ratio, gradually changes
in time?

\citet{colombo1966} was the first to point out the importance of the above
simple model in the study of the obliquity (the angle between the spin and
orbital axes) of planets and satellites. Subsequent works
\citep{peale1969,peale1974possible,ward1975tidal,henrard1987} have revealed rich
dynamics of this model. With appropriate modification, this model can be used as
a basis for understanding the evolution of rotation axes of celestial bodies.
Indeed, many contemporary problems in planetary/exoplanetary dynamics can be
cast into a form analogous to this simple model or its variants
\citep[e.g.][]{ward2004I, fabrycky_otides, batygin2013magnetic, lai2014star,
anderson2018teeter,zanazzi2018planet}.

In this paper we present a systematic investigation on the secular evolution of
Colombo's top, starting from general initial conditions. Our study includes
several new analytical results that go beyond previous works. While our results
are general, we frame our study in the context of generating exoplanet
obliquities from planet-disk interaction with a dissipating disk.

@@

The Millholland \& Batygin study assumes a negligible initial planetary
obliquity. This assumption is intuitive, since the planet attains its spin
angular momentum from the disk. But it may not always be satisfied. In
particular, the formation of rocky planets through planetesimal accretion can
lead to a wide range of obliquities, especially if the final spin is imparted by
a few large bodies \citep{dones1993does, lissauer1997accretion,
miguel2010planet}. Such ``stochastic'' accretion likely happened for terrestrial
planets in the Solar System. Giant impacts may have also played a role in the
formation of the close-in multiple-planet systems diskovered by the Kepler
satellite \citep[e.g.][]{inamdar2015formation, izidoro2017breaking}.

@@

The paper is organized as follows. In Section~\ref{s:eq}, we review the relevant
spin-orbit dynamics and key concepts that are used in the remainder of the
paper. In Sections~\ref{s:ad} and~\ref{s:nonad}, we study the evolution of the
system when the disk dissipates on different timescales, from highly adiabatic
to nonadiabatic. Analytical results are presented to explain the numerical
results in both limits. We diskuss the implications of our results in
Section~\ref{s:disk}. Our primary physical results consist of
Fig.~\ref{fig:ad_ensemble} in the adiabatic limit and
Fig.~\ref{fig:3_ensemble_05_15} in the nonadiabatic limit. Some detailed
calculations are relegated to the appendices, including a leading-order
estimate of the final planetary obliquities given small initial spin-disk
misalignment angles in Appendix~\ref{s:ad_approx}.

@@

The spin axis $\uv{s}$ of the planet tends to precess around its orbital
(angular momentum) axis $\uv{l}$, driven by the gravitational torque from the
host star acting on the planet's rotational bulge. On the other hand, $\uv{l}$
and the disk axis $\uv{l}_{\rm d}$ precess around each other due to
gravitational interactions. We assume $S \ll L \ll L_{\rm d}$, so $\uv{l}_{\rm
d}$ is nearly constant and $\uv{l}$ experiences negligible backreaction torque
from $\uv{s}$. The
equations of motion for $\uv{s}$ and $\uv{l}$ are
\begin{align}
    \rd{\uv{s}}{t} &= \omega_{\rm sl} \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        \equiv \alpha \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}},\label{eq:dsdt}\\
    \rd{\uv{l}}{t} &= \omega_{\rm ld}\p{\uv{l} \cdot \uv{l}_{\rm d}}
            \p{\uv{l} \times \uv{l}_{\rm d}}
        \equiv -g\p{\uv{l} \times \uv{l}_{\rm d}},
            \label{eq:dldt}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &\equiv \frac{3GJ_2 M_pR_p^2 M_\star}{2a_p^3 I_p\Omega_p}
        = \frac{3k_{\rm qp}}{2k_{\rm p}} \frac{M_\star}{m_{\rm p}}
            \p{\frac{R_{\rm p}}{a_{\rm p}}}^3 \Omega_{\rm p},\label{eq:wsl}\\
    \omega_{\rm ld} &\equiv \frac{3M_{\rm d}}{4M_\star}\p{\frac{a_{\rm
            p}}{r_{\rm d}}}^3 n .\label{eq:wld}
\end{align}
In Eq.~\eqref{eq:wsl}, $I_p = k_p M_pR_p^2$ (with $k_p$ a constant) is the
moment of inertia and $J_2=k_{\rm qp}\Omega_p^2 (R_p^3/GM_p)$ (with $k_{qp}$ a
constant) the rotation-induced (dimensionless) quadrupole of the planet [for a
body with uniform density, $k_p=0.4, k_{qp}=0.5$; for giant planets, $k_p\simeq
0.25$ and $k_{qp}\simeq 0.17$ \citep[e.g.][]{lainey2016quantification}]. In
Eq.~\eqref{eq:wld}, $n \equiv \sqrt{GM_\star/a_{\rm p}^3}$ is the planet's
orbital mean motion, and we have assumed $r_d\gg a_p$ and included only the
leading-order (quadrupole) interaction between the planet and disk. We define
three relative inclination angles via
\begin{equation}
  \uv{s} \cdot \uv{l}\equiv \cos\theta,\quad
  \uv{s} \cdot \uv{l}_{\rm d}\equiv \cos\theta_{\rm sd},\quad
  \uv{l} \cdot \uv{l}_{\rm d}\equiv \cos I.
\end{equation}
In our model, $I$ is a constant.
Following standard notation \citep[e.g.][]{colombo1966,peale1969,ward2004I}, we
have defined $\alpha \equiv \omega_{\rm sl}$ and $g \equiv -\omega_{\rm ld}\cos I$.

@@

In this section, we study the evolution of the planetary obliquity $\theta$ when
the parameter $\eta$ [or the disk mass $M_{\rm d}$; see Eqs.~\eqref{eq:eta}
and~\eqref{eq:deta_dt}] decreases sufficiently slowly that the evolution is
adiabatic. Intuitively, this requires the disk evolution time
$t_{\rm d}$ [Eq.~\eqref{eq:dmd_dt}] be much larger than the spin precession
period, $2\pi / \alpha$, i.e.\ $\epsilon = 1 / (\alpha t_{\rm d}) \ll 1/
(2\pi)$.

More rigorously, adiabaticity requires $t_{\rm d}$ be much larger than all
timescales of the dynamical system governed by the Hamiltonian
[Eq.~\eqref{eq:H}]. This is of course not possible in all cases, as the motion
along the separatrix has an infinite period. In practice, as $\eta$ evolves, the
system only crosses the separatrix once or twice, while it spends many orbits
inside one of the three zones and far from the separatrix. Thus, a \emph{weak
adiabaticity criterion} is that, for all equilibria/fixed points, the local
circulation/libration periods are much shorter than the timescale for the motion
of the equilibria due to changing $\eta$. If this criterion is satisfied, then
the system will evolve adiabatically for most of its evolution save one or two
separatrix crossings.

As shown in Appendix~\ref{ss:eigens}, libration about CS2 is slower than that
about CS1 or CS3. As such, it has the smallest characteristic frequency in the
system. The weak adiabaticity criterion is equivalent to
requiring that, at all times other than separatrix crossing, the obliquity of
CS2 ($\theta_2$) evolve over a longer timescale than the local libration period
about CS2, i.e.
\begin{equation}
    \abs{\rd{\theta_2}{\tau}} \ll
        \frac{\omega_{\rm lib}}{2\pi},\label{eq:ad_base}
\end{equation}
where
\begin{equation}
    \omega_{\rm lib} = \sqrt{\eta\sin I \sin \theta_2
        \p{1 + \eta \sin I \csc^3 \theta_2}},\label{eq:w_lib}
\end{equation}
is the libration frequency about CS2 for a given $\eta$
(Appendix~\ref{ss:eigens}). This formula differs from that given in
\citet{millholland_disk}, where the $\csc^3\!\theta_2$ term is neglected and the
square root is missing\footnote{The missing $\csc^3\!\theta_2$ term can be
traced to a $\theta \gg I$ approximation made in Eq.~(3) of \citet{ward2004II}.
Since $\theta_2 \sim I$ for $\eta \gg 1$ (Fig.~\ref{fig:cs_locs}), this
approximation is not always valid.}. Differentiating Eq.~\eqref{eq:cs_zero}
gives
\begin{equation}
    \rd{\theta_2}{\tau} = -\epsilon \frac{\eta \sin\p{\theta_2 - I}}{
        \cos\p{2\theta_2} - \eta \cos\p{\theta_2 - I}},\label{eq:dq2dt}
\end{equation}
where $\epsilon = -\rdil{(\ln \eta)}{\tau}$ [Eq.~\eqref{eq:eps_def}].
Eq.~\eqref{eq:ad_base} is most constraining at $\eta \sim 1$, i.e.\ it will be
satisfied for all $\eta$ if it is satisfied near $\eta \sim 1$, where
$\abs{\rdil{\theta_2}{\tau}} \sim \epsilon$. Thus, weak adiabiticity requires
\begin{equation}
    \epsilon \ll \epsilon_{\rm c}
        \equiv \p{\frac{\omega_{\rm lib}}{2\pi}}_{\eta = 1}
            \simeq \frac{1}{2\pi\sqrt{2}}\sqrt{\sin I\p{1 + 8\sin I}},
            % =\frac{1}{2\pi}\sqrt{\sin I \sin \theta_{\rm 2c}
            %     \p{1 + \sin I \csc^3 \theta_{\rm 2c}}}
            % \approx 0.0433,
            \label{eq:ad_constr}
\end{equation}
where in the last equality we have used $\sin \theta_{\rm 2} \simeq 1/2$ at
$\eta = 1$ (e.g.\ when $I = 5^\circ$ and $\eta = 1$, $\theta_{\rm 2} \approx
31^\circ$). For $I = 5^\circ$, we obtain $\epsilon_{\rm c} \approx 0.0433$.
Since our criterion is only a weak condition for adiabaticity, we use $\epsilon
= 3 \times 10^{-4}$ in our ``adiabatic'' calculations below. We explore the
consequences of nonadiabatic evolution in Section~\ref{s:nonad}.

@@

To obtain the distribution of the final obliquities $\theta_{\rm f}$, we evenly
sample $101$ values of $\theta_{\rm sd, i}$, and for each $\theta_{\rm sd, i}$
value, we pick $101$ evenly spaced orientations of $\uv{s}$ approximately from
the ring of initial conditions having angular distance $\theta_{\rm sd, i}$ to
$\uv{l}_{\rm d}$%
%
\footnote{The actual procedure we adopt to choose the initial conditions is the
natural extension of this description to finite $\eta_{\rm i}$. Note that the
center of libration of $\uv{s}$ is CS2, which, since $\eta_{\rm i}$ is finite,
is different from $\uv{l}_{\rm d}$. Furthermore, the libration is not exactly
circular. As a result, the libration trajectories for initial conditions on the
circular ring of points having angular distance $\theta_{\rm sd, i}$ from
$\uv{l}_{\rm d}$ are not the same and will each enclose slightly different
initial phase space areas $A_{\rm i}$. Since our analytical theory assumes exact
conservation of the initially enclosed phase space area $A_{\rm i}$ for each
$\theta_{\rm sd, i}$ (see Section~\ref{ss:zone_transitions}), this diskrepancy
introduces an extra deviation from the analytical prediction. To guarantee all
points for a particular $\theta_{\rm sd, i}$ have the same $A_{\rm i}$, we
instead choose initial conditions on the libration cycle going through
$\p{\theta_2 + \theta_{\rm sd, i}, \phi_2}$ [where $\p{\theta_2, \phi_2}$ are
the coordinates of CS2]. This ensures that all initial conditions for a given
$\theta_{\rm sd, i}$ enclose the same initial $A_{\rm i}$. As $\eta_{\rm i} \to
\infty$, this procedure generates initial conditions on the ring having angular
distance $\theta_{\rm sd, i}$ to $\uv{l}_{\rm d}$, recovering the procedure
given in the text.}.
%
To be concrete, we choose $\eta_{\rm i} = 10\eta_{\rm c}$ where $\eta_{\rm c}$
is given by Eq.~\eqref{eq:etac} and evolve Eqs.~\eqref{eq:dsdt_base}
and~\eqref{eq:deta_dt} until $\eta$ reaches its final value $10^{-5}$. At such a
small $\eta$, $\uv{s}$ is strongly coupled to $\uv{l}$ and the final obliquity
$\theta_{\rm f}$ is frozen. The mapping between $\theta_{\rm sd, i}$ and
$\theta_{\rm f}$ is our primary result, and is shown for $I = 5^\circ$,
$10^\circ$, and $20^\circ$ in
Figs.~\ref{fig:ad_ensemble},~\ref{fig:3_ensemble_10_35},
and~\ref{fig:3_ensemble_20_35} respectively. The blue dots represent the results
of the numerical calculation. The colored tracks are calculated
semi-analytically using the method diskussed in the following subsection.

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. We adopt
convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
This definition of $A$ has two advantages: (i) it is continuous across
transitions from circulating to librating that cross the North pole ($\cos
\theta = 1$), and (ii) the areas of the three zones are equal in absolute value
to the expressions given in Eqs.~\eqref{se:area_ward}. The path over which the
integral is taken is either a libration or circulation cycle. When $\eta_{\rm i}
\gg 1$, trajectories librate about $\uv{l}_{\rm d}$ with constant $\theta_{\rm
sd}$, meaning they enclose initial phase space area
\begin{equation}
    A_{\rm i} = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_{\rm i}$, as trajectories near
CS2 or CS3 librate about these equilibria, rather than $\uv{l}_{\rm d}$, and
Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice, Eq.~\eqref{eq:ai_qsd}
holds very well when defining $\theta_{\rm sd, i}$ as the angular distance to
CS2; an exception is diskussed in Section~\ref{sss:evol_traj}.

@@

The above calculation applies for a specific initial $\theta_{\rm i}$, but, as
diskussed at the beginning of Section~\ref{ss:ad_ensemble}, the initial spin
orientation is more appropriately described by $\theta_{\rm sd, i}$ since
$\eta_i\gg 1$. The correct way to predict the final obliquity for a given
$\theta_{\rm sd, i}$ using Eq.~\eqref{eq:thetaf} is somewhat subtle but yields
good agreement with numerical results.

@@

We thank Kassandra Anderson for diskussion and assistance in the early phase of
this work. DL thanks the Dept.\ of Astronomy and the Miller Institute for Basic
Science at UC Berkeley for hospitality while part of this work was carried out.
This work has been supported in part by the NSF grant AST-17152 and NASA grant
80NSSC19K0444. YS is supported by the NASA FINESST grant
19-ASTRO19-0041.% chktex 8
